---
title: "The party problem with datree"
author: "Jongbin Jung"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The party problem with datree}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

As a starting point, we can use `datree` to solve Kim's party problem, from the
Decision Analysis courses at Stanford or the Foundations of Decision Analysis 
textbook by Howard and Abbas.

```{r tree definition}
library(datree)

# Create outcomes
outdoor_sun <- o_node(~ v_os, "Outdoor-Sun")
porch_sun <- o_node(~ v_ps, "Porch-Sun")
indoor_rain <- o_node(~ v_ir, "Indoor-Rain")
indoor_sun <- o_node(~ v_is, "Indoor-Sun")
porch_rain <- o_node(~ v_pr, "Porch-Rain")
outdoor_rain <- o_node(~ v_or, "Outdoor-Rain")

# Uncertainty nodes at each alternative
outdoor <- u_node(list(outdoor_sun, outdoor_rain),
                  probs = c(~ p_sun, ~ 1- p_sun),
                  name = "Outdoor")
porch <- u_node(list(porch_sun, porch_rain),
                  probs = c(~ p_sun, ~ 1- p_sun),
                  name = "Porch")
indoor <- u_node(list(indoor_sun, indoor_rain),
                  probs = c(~ p_sun, ~ 1- p_sun),
                  name = "Indoor")

# Decision of party location
location <- d_node(list(outdoor, porch, indoor),
                   "Location")
```

The nodes of the tree are defined with abstract values --- expressed with
formula (starting with `~`).
Actual variable values must be defined within the calling environment:

```{r initial values}

# Load initial values
v_os <- 1
v_ps <- .95
v_ir <- .67
v_is <- .57
v_pr <- .32
v_or <- 0
p_sun <- .4
```

Once the variables are defined, you can compute the concrete value of any node
for the given parameters, using `evaluate()`.

```{r}
evaluate(outdoor)
evaluate(indoor)
evaluate(porch)

evaluate(location)
```

Note that for sensitivity analysis, you can also provide a `list`-like data type (e.g., `data.frame`)
that replaces the global environment values.

```{r}
evaluate(outdoor, list(p_sun = .5))
```

Also, for now, it's up to *you* to keep track and make sure the probabilities add to one!

```{r}
# This makes no sense, but still works ... (unfortunately)
evaluate(outdoor, list(p_sun = 500))
```

Use `decide()` to explore the different alternatives of a decision node.

```{r}
decide(location)
```

Like `evaluate()`, you can also specify parameters for `decide()`.

```{r}
decide(location, list(p_sun = .8))
```

This setup works well for sensitivity analysis.

```{r}
library(tidyverse)
map_df(c(0, .5, 1), ~ cbind(
  decide(location, list(p_sun = .x)), p_sun = .x)) %>% 
  ggplot(aes(x = p_sun, y = evalue)) +
  geom_line(aes(color = alternative))
```

Alternatively, the tree can be built as recursive calls to node functions, which
in some situations might be more readable, especially if you don't need/want to
keep track of intermediate nodes.

```{r}
# Decision of party location
location <- d_node(list( # Alternatives
  # Outdoor
  u_node(list(  # Outcomes for Outdoor
    o_node(~ v_os, "Outdoor-Sun"),
    o_node(~ v_or, "Outdoor-Rain")), 
    probs = c(~ p_sun, ~ 1 - p_sun),  
    name = "Outdoor"),
  # Porch
  u_node(list(  # Outcomes for Porch
    o_node(~ v_ps, "Porch-Sun"),
    o_node(~ v_pr, "Porch-Rain")), 
    probs = c(~ p_sun, ~ 1 - p_sun),  
    name = "Porch"),
  # Indoor
  u_node(list(  # Outcomes for Indoor
    o_node(~ v_is, "Indoor-Sun"),
    o_node(~ v_ir, "Indoor-Rain")), 
    probs = c(~ p_sun, ~ 1 - p_sun),  
    name = "Indoor")
  ), "Location")

decide(location)
```


